---
title: "Nowcasting electricity access rates"
author: "Craig Jolley"
date: "November 14, 2016"
output: html_document
---

This document explains our methodology for "nowcasting" electricity access rates based on publicly-available historical data that is often several years old. 

```{r, warning=FALSE, message=FALSE}
library(ggplot2)
library(dplyr)
library(llamar)
library(printr)
```

First, let's take a look at electricity access rates for a select group of African countries:

```{r}
dhs_elect <- loadDHS(indicators='HC_ELEC_H_ELC',
                     countries='ET,GH,KE,LB,MW,NG,RW,SN,SL,TZ,UG,ZM') %>%
  mutate(elect=Value) %>%
  dplyr::select(CountryName,SurveyYear,elect) 

dhs_elect %>% 
  mutate(elect=elect/100) %>%
  ggplot(aes(x=SurveyYear,y=elect,group=CountryName,color=CountryName)) +
    geom_point(size=4) +
    geom_line(size=2) +
    xlab('Year') +
    ylab('Electricity access') + 
    theme_classic() +
    scale_y_continuous(labels = scales::percent) +
    theme(axis.ticks=element_blank(),
          legend.title=element_blank())
```

You'll notice that, while access in increasing in most places, a lot of countries don't have particularly recent data.

Our first attempt will be a linear model with a constant growth rate for all countries and constant offsets for each.

```{r}
fit <- lm(elect ~ CountryName + SurveyYear,data=dhs_elect)
summary(fit)
```

The really important row there is ```SurveyYear``` -- all else being equal, the access rate for African countries increases by about 0.92% per year. Let's add this to ```dhs_elect``` and update the visualization.

```{r}
newpts <- data.frame(CountryName=unique(dhs_elect$CountryName),
                     SurveyYear=2016)
newpts <- newpts %>%
  mutate(elect=predict(fit,newpts))

rbind(dhs_elect,newpts) %>%
  ggplot(aes(x=SurveyYear,y=elect,group=CountryName,color=CountryName)) +
    geom_point(size=4) +
    geom_line(size=2) +
    theme_classic()
```

Not so impressive. Rapidly-electrifying countries like Ghana, Senegal, and Kenya appear to be dramatically underestimated, because the average rate is dominated by the more slowly-electrifying countries. We are probably better off choosing a model that allows a different growth rate for each country.

```{r}
elect_pred <- function(country,year=2016) {
  dhs_elect %>% filter(CountryName==country) %>%
    lm(elect ~ SurveyYear,data=.) %>%
    predict(data.frame(SurveyYear=year))
}
  
newpts <- data.frame(CountryName=unique(dhs_elect$CountryName),
                     SurveyYear=2016)
for (n in newpts$CountryName) {
  newpts[newpts$CountryName==n,'elect'] <- elect_pred(n)
}

rbind(dhs_elect,newpts) %>% 
  ggplot(aes(x=SurveyYear,y=elect,group=CountryName,color=CountryName)) +
  geom_point(size=4) +
  geom_line(size=2) +
  theme_classic()
```
 
This model is more complex, but produces much more reasonable-looking estimates for countries with rapid electrification. Kenya, for example, still shows a drop in 2016. The linear model assumes, based on Kenya's historical trajectory, that the rapid increase from 36% in 2014 to 41.6% in 2015 was a fluke, and that 2016's value should be a more modest 38.2%. A non-linear model might see the 2014-5 increase as the beginning of an exponential takeoff, and runs the risk of producing overly-exuberant estimates of 2016 electrification rates.

Now that we have country-level estimates, we might inquire into which individual households were most likely to have acquired electricity access between the last measured data point and the present. The DHS survey reports a composite index of household wealth -- is there a strong relationship between wealth and electricity access?

```{r warning=FALSE}
source('read-DHS.R')
wealth_elect <- function(df,text) {
  df %>% removeAttributes %>%
    ggplot(aes(x=wealth,y=elect)) +
    geom_jitter(size=5,color='tomato',alpha=0.05,width=0) +
    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
    ggtitle(text) +
    xlab('Asset-based wealth index') +
    ylab('Electricity access (0=no, 1=yes)') +
    theme_classic()
}

wealth_elect(rwa_2015,'Rwanda 2015')
```

In Rwanda (with sort of middle-of-the-pack electrification rates), there seems to be a sigmoidal relationship. Households with a low wealth index (the majority) are unlikely to have electric access, while wealther households are more likely to be electrified.

```{r}
wealth_elect(gha_2014,'Ghana 2014')
```

A similar relationship exists in Ghana, where overall electrification rates are much higher. Notice that this isn't just because Ghanians are wealthier -- the wealth index is designed to be comparable across countries, so people with the same overall level of wealth are more likely to have access to electricity in Ghana than in Rwanda.

If electricity access depends on wealth, then is electricity also getting more affordable with time? In other words, is it likely that people with the same overall level of wealth are able to afford electricity at later times than at earlier ones?

```{r warning=FALSE}
p50 <- function(df) {
  s <- df %>% removeAttributes %>%
    glm(elect~wealth,data=.,family='binomial') %>%
    summary
  data.frame(res=-s$coefficients[1,1]/s$coefficients[2,1])
}
elect_compare <- function(a,title='') {
  l <- a$label %>% unique
  # if (length(l)==2) {
  #   p1 <- a[a$label==l[1],] %>% p50
  #   p2 <- a[a$label==l[2],] %>% p50
  #   print(paste('p50 difference = ',round(p2-p1,digits=3)))
  # }
  ggplot(a,aes(x=wealth,y=elect,group=label,color=label)) +
    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
    ggtitle(title) +
    xlab('Asset-based wealth index') +
    ylab('Probability of electricity access') +
    theme_classic() +
    theme(legend.title=element_blank(),
          legend.background=element_blank())
}

rbind(gha_2008,gha_2014) %>% removeAttributes %>%
  elect_compare(title='Ghana 2008-2014') 
```

In Ghana, somewhat less wealthy people had access to electricity in 2014 than in 2008.

```{r warning=FALSE}
rbind(rwa_2010,rwa_2015) %>% removeAttributes %>%
  elect_compare(title='Rwanda 2010-2015')
```

Rwanda actually has the largest difference of any of the countries we looked at; electricity came within reach of many more households in 2015 than in 2010.

Wealth isn't the only factor -- affordable electricity access also depends on the presence of grid infrastructure, which is going to depend on geography. If your neighbors have electricity, it seems more likely that you will. DHS separates responses into geographic clusters, so we can test this hypothesis by determining whether a household is more likely to have access when their neighbors do.

```{r warning=FALSE}
geo_est <- function(df,text=NULL) {
  mod<- df %>% group_by(clust) %>% 
    summarise(clust_avg=mean(elect,na.rm=TRUE),in_clust=n()) %>%
    plyr::join(df,by='clust') %>%
    mutate(out_avg = clust_avg*in_clust/(in_clust-1) - as.numeric(elect)/(in_clust-1)) %>%
    select(-clust_avg,-in_clust)
  mod %>% removeAttributes %>%
    ggplot(aes(x=out_avg,y=elect)) +
    geom_jitter(size=5,color='darkgoldenrod3',alpha=0.05,width=0) +
    geom_smooth(method = "glm", method.args = list(family = "binomial")) +
    ggtitle(text) +
    xlab('Fraction of neighbors with electricity') +
    ylab('Electricity access (0=no, 1=yes)') +
    theme_classic()  
}

geo_est(rwa_2015,'Rwanda 2015') 
```

While the geographic trend isn't as strong as the wealth trend, people with electricity are more likely to have neighbors with electricity. Given that wealth is also geographically-correlated, we can combine these factors in a linear model.

```{r}
compare_models <- function(df) {
  print("Wealth only")
  df %>% removeAttributes %>% 
    glm(data=.,elect ~ wealth,family='binomial') %>% summary %>% print
  mod <- df %>% group_by(clust) %>% 
    summarise(clust_avg=mean(elect,na.rm=TRUE),in_clust=n()) %>%
    plyr::join(df,by='clust') %>%
    mutate(out_avg = clust_avg*in_clust/(in_clust-1) - as.numeric(elect)/(in_clust-1)) %>%
    select(-clust_avg,-in_clust)
  print("Geography only")
  mod %>% removeAttributes %>% 
    glm(data=.,elect ~ out_avg,family='binomial') %>% summary %>% print
  print("Wealth and geography")
  mod %>% removeAttributes %>% 
    glm(data=.,elect ~ wealth+out_avg,family='binomial') %>% summary %>% print
}

compare_models(rwa_2015)
```

As expected, wealth does a significantly better job of explaining electrification than geography (AIC=6869.6 vs. 8047.4), but combining the two yields an even better result (AIC=5926.4). Although geography and wealth are correlated, both have highly significant coefficients when used together, suggesting that the best predictive approach will be to combine both.

The predictions in ```nowcast-electric.R``` therefore assume that:

* The probability of electrification can be predicted from wealth and geography
* Over time, electricity becomes accessible to people with lower wealth
* Current (or future) electrification rates at a national level can be estimated from historical trends.

Our approach, therefore, is to model the probability of acquiring electricity for each household in the DHS dataset at the time of the survey. We then estimate the overall national access rate, and assume that households are electrified in decreasing order of probability until the (properly-weighted) average is equal to the projected national access rate. This (combined with similar calculations for mobile ownership) is what allows us to estimate the current fraction of households with access to mobile phones but not electricity (see ```latent_demand.R```).